---
title: "Draft: Results"
---

## Study Sample Characteristics and Descriptive Statistics

The analysis uses two datasets, water temperature observations from NOAA station DPHA1 near Dauphin Island, Alabama, and electricity demand data from the U.S. Energy Information Administration (EIA).

## Water Temperature Data

The water temperature dataset has hourly measurements of water temperature (WTMP) measured in degrees Celsius. After cleaning the data and removing missing values we are left with data spanning 2025. Water temperature values were converted from Celsius to Fahrenheit to correspond with the 85°F threshold.

```{r, include=FALSE}
library(tidyverse)
library(lubridate)

temp <- read_table(
  "water_temp.txt",
  comment = "#"
)

temp_clean <- temp %>%
  mutate(
    YY = as.integer(YY),
    MM = as.integer(MM),
    DD = as.integer(DD),
    hh = as.integer(hh),
    mm = as.integer(mm),
    WTMP = as.numeric(WTMP)
  ) %>%
  filter(
    !is.na(WTMP),
    WTMP > -5,
    WTMP < 40
  ) %>%
  mutate(
    datetime = lubridate::make_datetime(YY, MM, DD, hh, mm),
    temp_f = WTMP * 9/5 + 32
  )
```

The data shows that temperatures vary seasonally throughout the year. Lower temperatures occur during winter months and higher temperatures during summer. We see continuous examples of the water being above 85°F during the summer months.

## Electricity Demand Data

The electricity demand dataset is also measured hourly, deriving electricity demand values measured in megawatts (MW). Demand levels vary throughout the day and across seasons, but higher values are observed during summer and winter months when heating and cooling are in demand.

```{r, include=FALSE}
demand <- read_csv("energy.csv")

demand_clean <- demand %>%
  mutate(
    date = mdy(`Data Date`),
    hour = as.numeric(`Hour Number`) - 1,
    datetime = date + hours(hour),
    demand_mw = as.numeric(`Demand (MW)`)
  ) %>%
  filter(!is.na(datetime), !is.na(demand_mw))
```

## Primary Analysis

The primary analysis looks at instances of water temperatures exceeding 85°F. These values represent periods when Gulf water temperatures reach levels associated with higher heat conditions.

The analysis identifies the frequency and duration of observations where water temperature exceeds this threshold. This occurs mainly during summer months and appear in clusters of consecutive hours or days.

Figure 1 shows the time series of water temperature observations.

```{r}
ggplot(temp_clean, aes(datetime, temp_f)) +
  geom_line() +
  geom_hline(yintercept = 85, linetype = "dashed") +
  labs(
    title = "Water Temperature at Dauphin Island",
    x = "Time",
    y = "Temperature (°F)"
  )
```

## Secondary Analysis

The secondary analysis examines electricity demand patterns during periods when water temperatures exceed 85°F.

Dates with water temperatures above 85°F were identified and overlaid on the electricity demand time series. Red points in Figure 2 indicate days when water temperatures exceeded the 85°F threshold.

```{r}
high_temp_dates <- temp_clean %>%
  filter(temp_f > 85) %>%
  mutate(date = as.Date(datetime)) %>%
  distinct(date)

daily_demand <- demand_clean %>%
  mutate(date = as.Date(datetime)) %>%
  group_by(date) %>%
  summarize(daily_demand_mw = mean(demand_mw, na.rm = TRUE))

ggplot(daily_demand, aes(date, daily_demand_mw)) +
  geom_line(color = "blue") +
  geom_point(data = daily_demand %>% filter(date %in% high_temp_dates$date),
             aes(date, daily_demand_mw), color = "red", size = 1.5) +
  labs(
    title = "Daily Electricity Demand with High Water Temperature Days Highlighted",
    x = "Date",
    y = "Demand (MW)"
  )
```

High water temperature days occur primarily during the summer portion of the electricity demand curve. Figure 2 shows that many of these days coincide with periods of higher electricity demand, but interestingly show these days also occur during local minimums. 

## Note to add second half of 2025 later