---
title: "Evaluation of Dauphin Island Power Plant"
author: "Collin Henkel"
output-dir: docs
format:
  revealjs:
    theme: simple
---

## Why this matters

- Power plants rely on water for cooling  
- Efficiency drops at high temperatures  
- Demand peaks during extreme heat  

**Key constraint:**  
- Above **85°F**, cooling efficiency decreases  

**Question:**  
What happens when this occurs during peak demand?

---

## Objective

Assess risk that a that the power plant must reduce output during high electricity demand  

Key threshold: **85°F water temperature**

---

## Data Sources

- NOAA (water temperature)
- EIA (electricity demand)

- Frequent observations  
- Year: 2025  

---

## Approach

- Convert temperature to °F
- Remove missing and irregular data
- Identify periods > 85°F  
- Compare with electricity demand  

Focus: timing and overlap  

---

## Example of data

```{r, include=FALSE}
library(tidyverse)
library(lubridate)

temp <- read_table(
  "water_temp.txt",
  comment = "#"
)

temp_clean <- temp %>%
  mutate(
    YY = as.integer(YY),
    MM = as.integer(MM),
    DD = as.integer(DD),
    hh = as.integer(hh),
    mm = as.integer(mm),
    WTMP = as.numeric(WTMP)
  ) %>%
  filter(
    !is.na(WTMP),
    WTMP > -5,
    WTMP < 40
  ) %>%
  mutate(
    datetime = lubridate::make_datetime(YY, MM, DD, hh, mm),
    temp_f = WTMP * 9/5 + 32
  )
```
```{r}
knitr::kable(head(temp_clean[, c("datetime", "temp_f")]))
```

---

```{r}
library(ggplot2)

ggplot(temp_clean, aes(temp_f)) +
  geom_histogram(bins = 40) +
  geom_vline(xintercept = 85, linetype = "dashed") +
  labs(title = "Distribution of Water Temperature")
```

## Water Temperature Over Time


```{r, echo=FALSE}
#| label: fig-dauphin-island-temp
#| fig-cap: Daily maximum water temperature at Dauphin Island in 2025. The dashed line marks the 85°F operational threshold.

temp_daily <- temp_clean %>%
  mutate(date = as.Date(datetime)) %>%
  group_by(date) %>%
  summarize(temp = max(temp_f, na.rm = TRUE))

ggplot(temp_daily, aes(date, temp)) +
  geom_line() +
  geom_hline(yintercept = 85, linetype = "dashed") +
  labs(
    title = "Daily Maximum Water Temperature",
    x = "Date",
    y = "Temperature (°F)"
  ) +
  theme_minimal()
```

---

## Electricity Demand Over Time

```{r}
library(tidyverse)

demand1 <- read_csv("energy.csv")
demand2 <- read_csv("EIA930_BALANCE_2025_Jul_Dec.csv")

demand <- bind_rows(demand1, demand2)

demand_clean <- demand %>%
  mutate(
    datetime = as.POSIXct(`Local Time at End of Hour`, format="%m/%d/%Y %H:%M"),
    demand_mw = as.numeric(`Demand (MW)`)
  ) %>%
  filter(!is.na(datetime), !is.na(demand_mw), demand_mw > 0)

demand_daily <- demand_clean %>%
  mutate(date = as.Date(datetime)) %>%
  group_by(date) %>%
  summarize(demand = mean(demand_mw, na.rm = TRUE))

ggplot(demand_daily, aes(date, demand)) +
  geom_line() +
  labs(
    title = "Daily Average Electricity Demand (2025)",
    x = "Date",
    y = "Demand (MW)"
  ) +
  theme_minimal()
```

---

## Temperature vs Demand

```{r, echo=FALSE}
#| label: fig-temp-demand
#| fig-cap: Daily maximum water temperature (top) and average electricity demand (bottom). Shaded regions indicate days when water temperature exceeded 85°F.

#daily temp
daily_temp <- temp_clean %>%
  mutate(date = as.Date(datetime)) %>%
  group_by(date) %>%
  summarize(temp = max(temp_f, na.rm = TRUE))

#daily demand
daily_demand <- demand_clean %>%
  mutate(date = as.Date(datetime)) %>%
  group_by(date) %>%
  summarize(demand = mean(demand_mw, na.rm = TRUE))

#combine
plot_data <- left_join(daily_temp, daily_demand, by = "date")

#reshape for faceting
plot_long <- plot_data %>%
  pivot_longer(
    cols = c(temp, demand),
    names_to = "variable",
    values_to = "value"
  )

ggplot(plot_long, aes(date, value)) +

#highlight hot days
  geom_rect(
    data = plot_data %>% filter(temp > 85),
    aes(xmin = date, xmax = date + 1, ymin = -Inf, ymax = Inf),
    inherit.aes = FALSE,
    fill = "red",
    alpha = 0.1
  ) +

  geom_line() +

  facet_wrap(~variable, ncol = 1, scales = "free_y") +

  labs(
    title = "Water Temperature and Electricity Demand",
    x = "Date",
    y = NULL
  ) +

  theme_minimal()
```

---

## Key Findings

- ~20% of observations exceed 85°F  
- High temperatures occur in summer  
- These periods align with higher demand  

---

## What this means

- Reduced capacity during peak demand  
- Increased system strain  
- Lower operational flexibility  

---

## Recommendations

- Explore alternative cooling methods  
- Plan around summer risk  
- Consider demand management strategies  

---

## Conclusion

- High-temperature risk is **real and seasonal**  
- Overlaps with peak demand  
- Requires proactive planning  

---

## Thank You

Questions?